## R code for analysis of the Spirit Cave faunal data.

Cyler Conrad, Department of Anthropology, University of New Mexico, cylerc@unm.edu

This document contain R code to reproduce the plots and statistical analysis presented in 

> Conrad, C., Higham, C. and Eda, M. (in press) Paleoecology and Forager Subsistence Strategies During the Pleistocene-Holocene Transition: A Reinvestigation of the Zooarchaeological Assemblage from Spirit Cave, Mae Hong Son Province, Thailand. Asian Perspectives. 

All data required to perform the analyses can be found at the University of New Mexico digital electronic repository (Conrad 2015). The development version of this document can be found at https://github.com/cylerc/AP_SC

Details of the computational environment and software dependencies are listed at the end of this document. 

References: 
Conrad, C. 2015. Archaeological Databases for Spirit Cave, Mae Hong Son Province, Thailand  [dataset]. University of New Mexico. http://repository.unm.edu/handle/1928/26730

```{r setup}
# set the base directory for knitr to the directory above this one
library(knitr)
opts_knit$set(root.dir = '../')
```

```{r load_libraries, message = FALSE, warning = FALSE}
# see the output of sessionInfo() at the bottom for package version numbers
library(Bchron)
library(car)
```

```{r radiocarbon calibration}
#1. Calibrate Spirit Cave radiocarbon dates. 
dates <- read.csv("dates.csv",stringsAsFactors = FALSE, check.names = FALSE)
dates
#Ben-I want to calibrate this table of dates, but I can't figure out how to use the csv file to do it I've manually input them below
ages=BchronCalibrate(ages=c(3042,2995,7400,7905,8550,8520,11350,9180,10910),
                     ageSds=c(37,40,300,390,200,290,560,360,580),
                     positions=c(1,1,1,2,2,`2a`,`3/4`,4,4),
                     calCurves="intcal13")
#...but I get this error: Error in BchronCalibrate(ages = c(3042, 2995, 7400, 7905, 8550, 8520,  : ages and calCurves should be of same length
#Any thoughts on this?
summary(ages)
#Should I include a plot for the paper???
plot(ages,withDepths=TRUE)
```

```{r ratio NRSP-NISP calculation}
#2. Calculate ratios of NRSP/NISP per layer at SC. Analyze these values with Spearmans rhoto see if they are driven by sample size. 
sc <- read.csv("sc.csv", stringsAsFactors = FALSE, check.names = FALSE)
sc
str(sc)

#Okay, so I need to take the NISP totals (excluding rows 33-36,51-53) and column totals (NRPS) then divide...

NISP <- sc[sc$Taxon[1:33,37:50], -1]
#I'm having a problem with this last command...do you know how I can tell R to just select those rows for the analysis?
NRSP <- colSums(sc[ ,-1])
NRSP_NISP <-  NISP / NRSP
NRSP_NISP <- t(relative_abundance)
NRSP_NISP <- data_frame(`Layer` = row.names(NRSP/NISP), 
                                 `NRSP/NISP` = NRSP_NISP[,1])

#Then I need to take these ratios and use NRSP to run a spearmans rho analysis...

####Here is my cheating way without actually using the raw table...
#Ratio stands for NISP/NRSP
ratio <- read.table(text="
Ratio NRSP
1.38 44     
1.45 372
2.83 113
3.96 107
4.49 1766
", header=TRUE)

p.corr <-cor(ratio)
p.corr
p.corr.pval <- p.corr;
for (i1 in 1:ncol(ratio)) {
  for (i2 in 1:ncol(ratio)) {
    p.corr.pval[i1,i2] <- cor.test(ratio[, i1], ratio[, i2])$p.value
  }
}
p.corr.pval
s.corr <- cor(ratio, method="spearman")
s.corr
ratio
s.corr.pval <- p.corr;
for (i1 in 1:ncol(ratio)) {
  for (i2 in 1:ncol(ratio)) {
    s.corr.pval[i1,i2] <- cor.test(ratio[, i1], ratio[, i2], method="spearman")$p.value
  }
}
s.corr.pval
```

```{r linear regression analysis}
#3. Linear regression analysis of NISP to NTAXA, Cook's distance and leverage analysis of these values a and Spearmans rho test with sample size.

#3.1 NISP to NTAXA.
scn <- read.table(text="
NISP NTAXA
44 10     
372 29
113 11
107 20
1766 34
", header=TRUE)
scn
plot(scn$NISP, scn$NTAXA, main="Spirit Cave NISP-NTAXA", xlab="NISP", ylab="NTAXA")
abline(lm.nisp.ntaxa)
lm.nisp.ntaxa<-lm((NTAXA)~(NISP), data=scn)
lm.nisp.ntaxa
summary(lm.nisp.ntaxa)

#3.2 MNI to NTAXA.
scm <- read.table(text="
MNI NTAXA
10 10     
42 29
10 11
17 20
49 34
", header=TRUE)
scm
plot(scm$MNI, scm$NTAXA, main="Spirit Cave MNI-NTAXA", xlab="MNI", ylab="NTAXA")
abline(lm.mni.ntaxa)
lm.mni.ntaxa<-lm((NTAXA)~(MNI), data=scm)
lm.mni.ntaxa
summary(lm.mni.ntaxa)

#Ben-Emily suggested that I include these plots in the paper - but I'm not too sure...do you have any advice on this? She thought it might be nice to try and make one plot showing both regression lines, but I couldn't figure out how to do this! 

#3.3 NTAXA and sample size (nrs=NTAXA spearmans rho).
nrs<- read.table(text="
                   NRSP NTAXA 
                   44 10      
                   373 31 
                   113 13 
                   107 22 
                   1767 36 
                   ", header=TRUE)
nrs
p.corr <-cor(nrs)
p.corr
p.corr.pval <- p.corr;
for (i1 in 1:ncol(nrs)) {
  for (i2 in 1:ncol(nrs)) {
    p.corr.pval[i1,i2] <- cor.test(nrs[, i1], nrs[, i2])$p.value
  }
}
p.corr.pval
s.corr <- cor(nrs, method="spearman")
s.corr
nrs
s.corr.pval <- p.corr;
for (i1 in 1:ncol(nrs)) {
  for (i2 in 1:ncol(nrs)) {
    s.corr.pval[i1,i2] <- cor.test(nrs[, i1], nrs[, i2], method="spearman")$p.value
  }
}
s.corr.pval

#3.4 Plot Cook's distance and leverage.
par(mfrow=c(1,2))
plot(lm.nisp.ntaxa, which = c(4,6))
######Ben-do you know how to change the names of the points in the latter plots so that they reflect the SC layers (1,2,2a,3,4)? It automatically plots them as 1,2,3,4,5.
```

```{r chi-square analysis}
#4. Chi-square test between NISPs per layer at SC.
#4.1 Chi-square test between Layer 4 and 3.
scl4l3 <-
  matrix(c(2,2,1,7,22,3,189,1,5,56,1,10,4,7,1,21,2,1,0,0,
           3,0,2,0,2,0,44,125,131,0,1,4,2,14,8,1,2,1,7,8,3,1073,0,
           0,0,0,1,0,0,1,0,0,0,2,1,0,2,0,1,4,2,2,1,0,1,0,2,2,1,2,6,2,
           1,0,0,1,1,1,0,0,0,0,0,0,0,70),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29","30",
                      "31","32","33","34","35","36","37","38","39","40","41","42","43")))
scl4l3
chisq.summary <- chisq.test(scl4l3, correct=FALSE)
chisq.summary
#4.2 Chi-square test between Layer 3 and 2a.
scl3l2a <-
  matrix(c(0,1,0,0,1,0,2,1,0,2,1,4,2,2,1,1,2,2,1,2,6,2,0,1,1,1,1,70,0,
           2,0,9,2,7,8,1,0,1,2,0,0,0,0,0,0,0,0,0,5,1,0,54,0,0,0,0,0,13),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29")))
scl3l2a
chisq.summary <- chisq.test(scl3l2a, correct=FALSE)
chisq.summary
#4.3 Chi-square test between Layer 2a and 2.
scl2al2 <-
  matrix(c(2,9,2,7,0,8,1,0,0,1,2,0,0,0,0,0,0,0,0,0,0,0,0,5,1,54,0,0,0,0,8,0,13,
           0,0,21,104,14,23,4,2,3,1,4,2,1,1,1,2,1,2,1,4,2,4,4,3,6,0,1,2,1,1,51,43,63),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29","30",
                      "31","32","33")))
scl2al2
chisq.summary <- chisq.test(scl2al2, correct=FALSE)
chisq.summary
#4.4 Chi-square test between Layer 2 and 1.
scl2l1 <-
  matrix(c(21,104,14,23,4,2,3,1,4,2,1,1,1,2,1,2,1,4,2,4,4,3,6,0,0,0,0,1,2,1,1,51,43,63,
           10,0,0,0,0,0,0,0,0,3,0,0,0,0,0,0,0,0,0,0,0,2,0,1,2,4,1,1,0,0,0,10,0,10),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29","30",
                      "31", "32", "33", "34")))
scl2l1
chisq.summary <- chisq.test(scl2l1, correct=FALSE)
chisq.summary
```

```{r burnt bone spearmans rho analysis}
#5. Analysis of burnt bone NISPs and sample size using Spearmans rho.
burn <- read.table(text="
                  NISP NRSP
                  9 44     
                  72 373
                  16 113
                  30 107
                  194 1767
                  ", header=TRUE)
burn
p.corr <-cor(burn)
p.corr
p.corr.pval <- p.corr;
for (i1 in 1:ncol(burn)) {
  for (i2 in 1:ncol(burn)) {
    p.corr.pval[i1,i2] <- cor.test(burn[, i1], burn[, i2])$p.value
  }
}
p.corr.pval
s.corr <- cor(burn, method="spearman")
s.corr
burn
s.corr.pval <- p.corr;
for (i1 in 1:ncol(burn)) {
  for (i2 in 1:ncol(burn)) {
    s.corr.pval[i1,i2] <- cor.test(burn[, i1], burn[, i2], method="spearman")$p.value
  }
}
s.corr.pval
```

```{r session_info}
# Computational environment and software dependencies for this analysis.
sessionInfo()
```

LICENSE
 
The MIT License (MIT)

Copyright (c) 2015 Cyler Conrad

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in
all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
THE SOFTWARE.