## R code for analysis of the Spirit Cave faunal data.

Cyler Conrad, Department of Anthropology, University of New Mexico, cylerc@unm.edu

This document contain R code to reproduce the plots and statistical analysis presented in 

> Conrad, C., Higham, C. and Eda, M. (in press) Paleoecology and Forager Subsistence Strategies During the Pleistocene-Holocene Transition: A Reinvestigation of the Zooarchaeological Assemblage from Spirit Cave, Mae Hong Son Province, Thailand. Asian Perspectives. 

All data required to perform the analyses can be found at the University of New Mexico digital electronic repository (Conrad 2015). The development version of this document can be found at https://github.com/cylerc/AP_SC

Details of the computational environment and software dependencies are listed at the end of this document. 

References: 
Conrad, C. 2015. Archaeological Databases for Spirit Cave, Mae Hong Son Province, Thailand  [dataset]. University of New Mexico. http://repository.unm.edu/handle/1928/26730

```{r setup}
# set the base directory for knitr to the directory above this one
# install.packages(c("knitr", "Bchron", "car", "ggplot2", "reshape2", "dplyr"))
library(knitr)
opts_knit$set(root.dir = '../')
```

```{r load_libraries, message = FALSE, warning = FALSE}
# see the output of sessionInfo() at the bottom for package version numbers
library(Bchron)
library(car)
library(ggplot2)
library(reshape2)
library(dplyr)
library(viridis)
```

```{r radiocarbon calibration}
#1. Calibrate Spirit Cave radiocarbon dates. 
dates <- read.csv("dates.csv",stringsAsFactors = FALSE, check.names = FALSE)
dates
ages <- BchronCalibrate(ages = dates$Ages,
                     ageSds = dates$ageSds,
                     positions = c(1,1,1,2,2,"2a","3/4",4,4),
                     calCurves = rep("intcal13", length(dates$Ages)))
summary(ages)
# see all calibrated dates
par(mfrow=c(3, 3)) # set up a grid of 3x3 plots
plot(ages) # put all the plots on that grid

# or better, we can combine all the calibrated dates into a single plot, like this: 
ages_densities <- BchronDensity(ages = dates$Ages,
                     ageSds = dates$ageSds,
                     calCurves = rep("intcal13", length(dates$Ages)))
# this is the plot that should go in the paper
plot(ages_densities, xlab='Age (cal years BP)')
```

```{r ratio NRSP-NISP calculation}
#2. Calculate ratios of NRSP/NISP per layer at SC. Analyze these values with Spearmans rhoto see if they are driven by sample size. 
sc <- read.csv("sc.csv", stringsAsFactors = FALSE, check.names = FALSE)
sc
str(sc)

#Okay, so I need to take the NISP totals (excluding rows 33-36,51-53) and column totals (NRPS) then divide...

NISP <- sc[c(1:33,37:50), -1]
NRSP <- colSums(sc[ ,-1])
NRSP_NISP <-  NISP / NRSP
NRSP_NISP <- t(relative_abundance) # what is this?
#This was my attempt to emulate the code you helped write for the turkey JASR paper! It is not needed here. 
NRSP_NISP <- data_frame(Layer = row.names(NRSP/NISP), 
                        `NRSP/NISP` = NRSP_NISP[,1])
#Then I need to take these ratios and use NRSP to run a spearmans rho analysis...

####Here is my cheating way without actually using the raw table...
#Ratio stands for NISP/NRSP
ratio <- read.table(text="
Ratio NRSP
1.38 44     
1.45 372
2.83 113
3.96 107
4.49 1766
", header=TRUE)
```

#To measure sample size bias we calculate a generalized linear model between both NISP and NTAXA (taxonomic abundance) and MNI and NTAXA per layer (Grayson and Delpech 1998; Lyman 2008:149-152). If a significant (at 95%) relationship is present then this suggests that sample size biases occur equally throughout the assemblage. A non-significant relationship indicates biases are present between layers.
#Analysis of Cook’s distance and leverage also provides inference into potential issues with sample size (Chatterjee and Hadi 1986). Cook’s distance identifies which samples have the greatest influence in a regression analysis; leverage examines which samples are the furthest from predicted values. Lastly, we use chi-square to test for differences in taxonomic composition between layers at Spirit Cave based on NISP. If subsistence strategies remain the same at Spirit Cave throughout each layer then we expect no significant difference in the chi-square analysis. 

```{r linear regression analysis 1 }
#3. Linear regression analysis of NISP to NTAXA, Cook's distance and leverage analysis of these values a and Spearmans rho test with sample size.

#3.1 NISP to NTAXA.
scn <- read.table(text="
NISP NTAXA
44 10     
372 29
113 11
107 20
1766 34
", header=TRUE)

scn

ggplot(scn, aes(NISP, NTAXA)) +
  geom_point(size = 3) +
  geom_smooth(method = "glm") +
  theme_minimal() +
  ggtitle("Spirit Cave NISP-NTAXA")

glm.nisp.ntaxa <- glm(NTAXA ~ NISP, data = scn, family = gaussian)
glm.nisp.ntaxa
summary(glm.nisp.ntaxa)
```

```{r linear regression analysis 2}
#3.2 MNI to NTAXA.
scm <- read.table(text="
MNI NTAXA
10 10     
42 29
10 11
17 20
49 34
", header=TRUE)

scm

ggplot(scm, aes(MNI, NTAXA)) +
  geom_point(size = 3) +
  geom_smooth(method = "glm") +
  theme_minimal() +
  ggtitle("Spirit Cave MNI-NTAXA")


glm.mni.ntaxa <- glm(NTAXA ~ MNI, data = scm, family = gaussian)
glm.mni.ntaxa
summary(glm.mni.ntaxa)
```

#Generalized linear modeling suggests that sample size bias is occurring at the site (t=2.258, p=0.04). There is a significant relationship between MNI and NTAXA (t=7.592, p=0.005) suggesting that as sample size increases, MNI also increases. NTAXA is also significantly driven by sample size (rs=0.9, p=0.02). Analysis of Cook’s distance and leverage indicates that layer 4 is influencing the regression relationship and is furthest from the predicted values (Figure 5). Given that layer 4 is most fragmented and has the most unidentifiable specimens, this suggests that bones from layer 4 closely represent the original sample excavated from Spirit Cave. Finally, chi-squared statistics show that every subsequent layer at Spirit Cave is significantly different (Table 5). Since sample size biases are present, these patterns are likely not driven by forager subsistence strategies.


```{r chi-square analysis}
#4.1 Chi-square test between Layer 4 and 3.
scl4l3 <-
  matrix(c(2,2,1,7,22,3,189,1,5,56,1,10,4,7,1,21,2,1,0,0,
           3,0,2,0,2,0,44,125,131,0,1,4,2,14,8,1,2,1,7,8,3,1073,0,
           0,0,0,1,0,0,1,0,0,0,2,1,0,2,0,1,4,2,2,1,0,1,0,2,2,1,2,6,2,
           1,0,0,1,1,1,0,0,0,0,0,0,0,70),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29","30",
                      "31","32","33","34","35","36","37","38","39","40","41","42","43")))
scl4l3
chisq.summary <- chisq.test(scl4l3, correct=FALSE)
chisq.summary
#4.2 Chi-square test between Layer 3 and 2a.
scl3l2a <-
  matrix(c(0,1,0,0,1,0,2,1,0,2,1,4,2,2,1,1,2,2,1,2,6,2,0,1,1,1,1,70,0,
           2,0,9,2,7,8,1,0,1,2,0,0,0,0,0,0,0,0,0,5,1,0,54,0,0,0,0,0,13),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29")))
scl3l2a
chisq.summary <- chisq.test(scl3l2a, correct=FALSE)
chisq.summary
#4.3 Chi-square test between Layer 2a and 2.
scl2al2 <-
  matrix(c(2,9,2,7,0,8,1,0,0,1,2,0,0,0,0,0,0,0,0,0,0,0,0,5,1,54,0,0,0,0,8,0,13,
           0,0,21,104,14,23,4,2,3,1,4,2,1,1,1,2,1,2,1,4,2,4,4,3,6,0,1,2,1,1,51,43,63),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29","30",
                      "31","32","33")))
scl2al2
chisq.summary <- chisq.test(scl2al2, correct=FALSE)
chisq.summary
#4.4 Chi-square test between Layer 2 and 1.
scl2l1 <-
  matrix(c(21,104,14,23,4,2,3,1,4,2,1,1,1,2,1,2,1,4,2,4,4,3,6,0,0,0,0,1,2,1,1,51,43,63,
           10,0,0,0,0,0,0,0,0,3,0,0,0,0,0,0,0,0,0,0,0,2,0,1,2,4,1,1,0,0,0,10,0,10),
         nrow = 2, byrow = TRUE,
         dimnames = list(
           "Layer" = c("4", "3"),
           "NISP" = c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15",
                      "16","17","18","19","20","21","22","23","24","25","26","27","28","29","30",
                      "31", "32", "33", "34")))
scl2l1
chisq.summary <- chisq.test(scl2l1, correct=FALSE)
chisq.summary
```

```{r burnt bone spearmans rho analysis}
#5. Analysis of burnt bone NISPs and sample size using Spearmans rho.
>>>>>>> upstream/master
burn <- read.table(text="
                  NISP NRSP
                  9 44     
                  72 373
                  16 113
                  30 107
                  194 1767
                  ", header=TRUE)
burn
p.corr <-cor(burn)
p.corr
p.corr.pval <- p.corr;
for (i1 in 1:ncol(burn)) {
  for (i2 in 1:ncol(burn)) {
    p.corr.pval[i1,i2] <- cor.test(burn[, i1], burn[, i2])$p.value
  }
}
p.corr.pval
s.corr <- cor(burn, method="spearman")
s.corr
burn
s.corr.pval <- p.corr;
for (i1 in 1:ncol(burn)) {
  for (i2 in 1:ncol(burn)) {
    s.corr.pval[i1,i2] <- cor.test(burn[, i1], burn[, i2], method="spearman")$p.value
  }
}
s.corr.pval
```


```{r session_info}
# Computational environment and software dependencies for this analysis.
sessionInfo()
```

LICENSE
 
The MIT License (MIT)

Copyright (c) 2015 Cyler Conrad

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in
all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
THE SOFTWARE.