## R code for analysis of the Spirit Cave faunal data.

Cyler Conrad, Department of Anthropology, University of New Mexico, cylerc@unm.edu

Ben Marwick, Department of Anthropology, University of Washington, bmarwick@uw.edu

This document contain R code to reproduce the plots and statistical analysis presented in 

> Conrad, C., Higham, C. and Eda, M. (in press) Paleoecology and Forager Subsistence Strategies During the Pleistocene-Holocene Transition: A Reinvestigation of the Zooarchaeological Assemblage from Spirit Cave, Mae Hong Son Province, Thailand. Asian Perspectives. 

All data required to perform the analyses can be found at the University of New Mexico digital electronic repository (Conrad 2015). The development version of this document can be found at https://github.com/cylerc/AP_SC

Details of the computational environment and software dependencies are listed at the end of this document. 

References: 
Conrad, C. 2015. Archaeological Databases for Spirit Cave, Mae Hong Son Province, Thailand  [dataset]. University of New Mexico. http://repository.unm.edu/handle/1928/26730

```{r setup}
# set the base directory for knitr to the directory above this one
# install.packages(c("knitr", "Bchron", "car", "ggplot2", "reshape2", "dplyr"))
library(knitr)
opts_knit$set(root.dir = '../', progress = FALSE)
```

```{r load_libraries, message = FALSE, warning = FALSE}
# see the output of sessionInfo() at the bottom for package version numbers
library(Bchron)
library(car)
library(ggplot2)
library(reshape2)
library(plyr)
library(dplyr)
library(viridis)
```

```{r radiocarbon calibration}
# 1. Calibrate Spirit Cave radiocarbon dates. 
dates <- read.csv("data/dates.csv", stringsAsFactors = FALSE, check.names = FALSE)
ages <- BchronCalibrate(ages = dates$Ages,
                     ageSds = dates$ageSds,
                     positions = c(1,1,1,2,2,"2a","3/4",4,4),
                     calCurves = rep("intcal13", length(dates$Ages)))
# see all calibrated dates
par(mfrow=c(3, 3)) # set up a grid of 3x3 plots
plot(ages) # put all the plots on that grid
```

```{r, results = 'hide'} 
# combine all the calibrated dates into a single plot (using results = 'hide' to hide the progress bar)
ages_densities <- BchronDensity(ages = dates$Ages,
                     ageSds = dates$ageSds,
                     calCurves = rep("intcal13", length(dates$Ages)))
```

```{r}
par(mfrow=c(1,1)) # reset so that the next plot is normal sized.
tiff("figures/Figure4.tif", res=600, compression = "lzw", height=5, width=5, units="in") # save as high resolution tif
plot(ages_densities, xlab='Age (cal years BP)')
dev.off()
```

```{r ratio NRSP-NISP calculation}
#2. Calculate ratios of NRSP/NISP per layer at SC. Analyze these values with Spearmans rho to see if they are driven by sample size. 
sc <- read.csv("data/sc.csv", stringsAsFactors = FALSE, check.names = FALSE)

# calculates NISPs per layer without the Mammalia, sm/md/lg/intermediate, Vertebrate sm/md, indeterminate and Indeterminate values, and column totals.

NISP <- sc[c(1:32,37:50), -1]
row.names(NISP) <- sc$Taxon[c(1:32,37:50)]

# compute NRSP (total number of specimens)

NRSP <- unname(colSums(sc[ ,-1]))

# and now the ratio of NRSP to NISP

NRSP_NISP <-  NRSP / colSums(NISP) 

# spearman's rho on the ratios to sample size
NRSP_NISP_cor <- cor.test(NRSP, unname(colSums(NISP)), method = 'spearman')
NRSP_NISP_cor_r <-  unname(NRSP_NISP_cor$estimate)
NRSP_NISP_cor_p <-  round(NRSP_NISP_cor$p.value, 3)
```

```{r linear regression analysis 1 }
#3. Linear regression analysis of NISP to NTAXA, Cook's distance and leverage analysis of these values and Spearmans rho test with sample size.

# compute NTAXA

NTAXA <- unname(colSums(sc!= 0))[-1]

# NISP to NTAXA

scn <-  data_frame(NISP = colSums(NISP),
                   NTAXA = NTAXA)

ggplot(scn, aes(NISP, NTAXA)) +
  geom_point(size = 3) +
  geom_smooth(method = "glm") +
  theme_minimal() +
  ggtitle("Spirit Cave NISP-NTAXA")

glm.nisp.ntaxa <- glm(NTAXA ~ NISP, data = scn, family = gaussian)
glm.nisp.ntaxa
summary(glm.nisp.ntaxa)
glm.nisp.ntaxa_summary <- summary(glm.nisp.ntaxa)
glm.nisp.ntaxa_summary_t <- round(glm.nisp.ntaxa_summary$coefficients[3], 3)
glm.nisp.ntaxa_summary_p <- round(glm.nisp.ntaxa_summary$coefficients[4], 3)

# cook's distance
tiff("figures/Figure6.tif", res=600, compression = "lzw", height=5, width=5, units="in") # save as high resolution tif
plot(glm.nisp.ntaxa, which=5) 
dev.off()

# r-squared values, from https://stat.ethz.ch/pipermail/r-help/2010-June/243113.html

# possibility 1
glm.nisp.ntaxa_summary_p_r2 <- cor(NTAXA, predict(glm.nisp.ntaxa))^2

# possibility 2
glm.nisp.ntaxa_summary_p_r2 <- 1 - (sum((NTAXA - predict(glm.nisp.ntaxa))^2) / sum((NTAXA - mean(NTAXA))^2))
```

```{r linear regression analysis 2}
#3.2 MNI to NTAXA.
MNI <- read.csv("data/MNI.csv", stringsAsFactors = FALSE, check.names = FALSE)
MNI <- colSums(MNI[,3:7], na.rm = TRUE)

scm <- data_frame(MNI = unname(MNI), 
                  NTAXA = NTAXA)

ggplot(scm, aes(MNI, NTAXA)) +
  geom_point(size = 3) +
  geom_smooth(method = "glm") +
  theme_minimal() +
  ggtitle("Spirit Cave MNI-NTAXA")

glm.mni.ntaxa <- glm(NTAXA ~ MNI, data = scm, family = gaussian)
glm.mni.ntaxa
summary(glm.mni.ntaxa)
glm.mni.ntaxa_summary <- summary(glm.mni.ntaxa)
glm.mni.ntaxa_summary_t <- round(glm.mni.ntaxa_summary$coefficients[3], 3)
glm.mni.ntaxa_summary_p <- round(glm.mni.ntaxa_summary$coefficients[4], 3)

# cook's distance
plot(glm.mni.ntaxa, which = 5)

# r-squared values, from https://stat.ethz.ch/pipermail/r-help/2010-June/243113.html

# possibility 1
glm.mni.ntaxa_r2 <- cor(NTAXA, predict(glm.mni.ntaxa))^2

# possibility 2
glm.mni.ntaxa_r2 <- 1 - (sum((NTAXA - predict(glm.mni.ntaxa))^2) / sum((NTAXA - mean(NTAXA))^2))
```

```{r chi-square analysis}
# 4.1 Chi-square test between Layer 4 and 3.

sc_layer_3_and_4 <-  as.data.frame((sc[, 5:6]))

# exclude taxa with no counts
sc_layer_3_and_4 <- sc_layer_3_and_4[!rowSums(sc_layer_3_and_4) == 0, ]

# compute chi-sq
sc_layer_3_and_4_chisq.summary <- chisq.test(t(sc_layer_3_and_4), correct = FALSE)

# 4.2 Chi-square test between Layer 3 and 2a.

sc_layer_3_and_2a <-  as.data.frame((sc[, 4:5]))

# exclude taxa with no counts
sc_layer_3_and_2a <- sc_layer_3_and_2a[!rowSums(sc_layer_3_and_2a) == 0, ]

# compute chi-sq
sc_layer_3_and_2a_chisq.summary <- chisq.test(t(sc_layer_3_and_2a), correct = FALSE)

# 4.3 Chi-square test between Layer 2a and 2.

sc_layer_2_and_2a <-  as.data.frame((sc[, 3:4]))

# exclude taxa with no counts
sc_layer_2_and_2a <- sc_layer_2_and_2a[!rowSums(sc_layer_2_and_2a) == 0, ]

# compute chi-sq
sc_layer_2_and_2a_chisq.summary <- chisq.test(t(sc_layer_2_and_2a), correct = FALSE)

# 4.4 Chi-square test between Layer 2 and 1.

sc_layer_1_and_2 <-  as.data.frame((sc[, 2:3]))

# exclude taxa with no counts
sc_layer_1_and_2 <- sc_layer_1_and_2[!rowSums(sc_layer_1_and_2) == 0, ]

# compute chi-sq
sc_layer_1_and_2_chisq.summary <- chisq.test(t(sc_layer_1_and_2), correct = FALSE)
```

```{r table of all chi-sq values}

# table of all chi-sq values
my_chi_sqs <- list(sc_layer_1_and_2_chisq.summary,  
                     sc_layer_2_and_2a_chisq.summary, 
                     sc_layer_3_and_2a_chisq.summary,
                     sc_layer_3_and_4_chisq.summary)

# extract chi-sq, df, and p-values into a table
my_chi_sqs_df <- ldply(my_chi_sqs, function(i) c(unname(i$statistic), unname(i$parameter),  i$p.value))
# add col names
names(my_chi_sqs_df) <- c('chi-sq', 'df', 'p')
# add row names
my_chi_sqs_df$layers <- c("1_and_2", "2_and_2a", "2a_and_3", "3_and_4")
# rearrange the cols to put names first
my_chi_sqs_df <- my_chi_sqs_df[,c(4,1:3)]
# show the table
my_chi_sqs_df
```

```{r burnt bone spearmans rho analysis}
#5. Analysis of burnt bone NISPs and sample size using Spearmans rho.

burn <- read.table(text="
                  NISP NRSP
                  9 44     
                  72 373
                  16 113
                  30 107
                  194 1767
                  ", header=TRUE)
burn
# spearman's rho on the ratios to sample size
burn_NRSP_NISP_cor <- cor.test(burn$NISP, burn$NRSP, method = 'spearman')
burn_NRSP_NISP_cor_cor_r <-  unname(burn_NRSP_NISP_cor$estimate)
burn_NRSP_NISP_cor_cor_p <-  round(burn_NRSP_NISP_cor$p.value, 3)
```

```{r session_info}
# Computational environment and software dependencies for this analysis.
sessionInfo()
```

LICENSE
 
The MIT License (MIT)

Copyright (c) 2015 Cyler Conrad & Ben Marwick

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in
all copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN
THE SOFTWARE.
